State of the Practice for Medical Imaging Software Based on Open Source Repositories

### 2024-06-10 StateOfPractice/Papers/SOP_MI_OSR/SOP_MI_OSR.tex
Edits to SOP for MI paper

\usepackage[round, sort]{natbib}

@@

@@

We aim to study the state of software development practice for Medical Imaging
(MI) software using data available in open source repositories.  MI tools use
images of the interior of the body (from sources such as Magnetic Resonance
Imaging (MRI), Computed Tomography (CT), Positron Emission Tomography (PET) and
Ultrasound) to provide information for diagnostic, analytic, and medical
applications \citep{FDA2021, enwiki:1034887445, Zhang2008}.
Figure~\ref{Fig_Example}, which shows an image of the brain, highlights the
importance and value of MI. Through MI medical practitioners and researchers can
noninvasively gain insights into the human body, including information on
injuries and illnesses. Given the importance of MI software and the high number
of competing software projects, we wish to understand the merits and drawbacks
of the current development processes, tools, and methodologies. We aim to assess
through a software engineering lens the quality of the existing software with
the hope of highlighting standout examples, and providing guidelines and
recommendations for future development.

@@

Not only do we wish to gain insight into the state of the practice for MI
software, we also wish to understand the impact of the often cited gap, or chasm,
between software engineering and research software \citep{Kelly2007,
Storer2017}. Although scientists spend a substantial proportion of their working
hours on software development \citep{Hannay2009, Prabhu2011}, many developers
learn software engineering skills by themselves or from their peers, instead of
from proper training \citep{Hannay2009}. \citet{Hannay2009} observe that many
scientists showed ignorance and indifference to standard software engineering
concepts. For instance, according to a survey by \citet{Prabhu2011}, more than
half of their 114 subjects did not use a proper debugger when coding.

To gain insight, we devised 10 research questions, which can be applied to MI,
as well as to other domains, of research software \citep{SmithEtAl2021,
SmithAndMichalski2022}.  We designed the questions to learn about the
community's interest in, and experience with, software artifacts, tools,
principles, processes, methodologies, and qualities.  When we mention artifacts
we mean the documents, scripts and code that constitutes a software development
project. Example artifacts include requirements, specifications, user manuals,
unit tests, system tests, usability tests, build scripts, API (Application
Programming Interface) documentation, READMEs, license documents, process
documents, and code.  Once we have learned what MI developers do, we then put
this information in context by contrasting MI software against the trends shown
by developers in other research software communities.

@@

To make the project feasible, we only cover MI visualization software.  As a
consequence we are excluding many other categories of MI software, including
Segmentation, Registration, Visualization, Enhancement, Quantification,
Simulation, plus MI archiving and telemedicine systems (Compression, Storage,
and Communication) (as summarized by \citet{Bankman2000} and
\citet{Angenent2006}).  We also exclude Statistical Analysis and Image-based
Physiological Modelling \citep{enwiki:1034877594} and Feature Extraction,
Classification, and Interpretation \citep{Kim2011}. Software that provides MI
support functions is also out of scope; therefore, we have not assessed the
toolkit libraries VTK \citep{SchroederEtAl2006} and ITK \citep{McCormick2014}.
Finally, Picture Archiving and Communication System (PACS), which helps users to
economically store and conveniently access images \citep{Choplin1992}, are
considered out of scope.

@@

We designed a general methodology to assess the state of the practice for
research software \citep{SmithEtAl2021, SmithAndMichalski2022}.  Our methodology
has been applied to MI software \citep{Dong2021} and Lattice Boltzmann Solvers
\citep{Michalski2021, SmithEtAl2024}.  This methodology builds off prior work to
assess the state of the practice for such domains as Geographic Information
Systems \citep{smith2018state}, Mesh Generators \citep{smith2016state},
Seismology software \citep{Smith2018Seismology}, and Statistical software for
psychology \citep{smith2018statistical}.  In keeping with the previous
methodology, we have maintained the constraint that the work load for measuring
a given domain should be feasible for a team as small as one person, and for a
short time, ideally around a person month of effort. We consider a person month
as $20$ working days ($4$ weeks in a month, with $5$ days of work per week) at
$8$ person hours per day, or $20 \times 8 = 160$ person hours.

With our methodology, we first choose a research software domain (in the current
case MI) and identify a list of about 30 software packages. (For measuring MI we
used 29 software packages.)  The selected packages are vetted by a domain
expert. We approximately measure the qualities of each package by filling in a
grading template. The grader looks at the contents of the open source repository
and repository based metrics, such as the number of files, number of lines of
code, percentage of issues that are closed, etc.  With the quantitative data in
the grading template, we rank the software with the Analytic Hierarchy Process
(AHP). Further details on interaction with the domain expert, software
qualities, grading the software and AHP are found below and in
\citep{SmithEtAl2024_MI_SOP}.

@@

Quality is defined as a measure of the excellence or worth of an entity.  As is
common practice, we do not think of quality as a single measure, but rather as a
set of measures.  That is, quality is a collection of different qualities, often
called ``ilities.''  For this study we selected 9 qualities to measure:
installability, correctness/ verifiability, reliability, robustness, usability,
maintainability, reusability, understandability, and visibility/transparency.
With the exception of installability, all the qualities are defined in
\citet{GhezziEtAl2003}. Installability is defined as the effort required for the
installation and/or uninstallation of software in a specified environment
\citep{ISO/IEC25010, lenhard2013measuring}.

@@

We grade the selected software using the measurement template summarized in
\citet{SmithEtAl2021}.  The template provides measures of the qualities listed
in Section~\ref{sec_software_quality}.
For each software package, we fill in the template questions. To stay within the
target of 160 person hours to measure the domain, we allocated between one and
four hours for each package. Project developers can be contacted for help
regarding installation, if necessary, but we impose a cap of about two hours on
the installation process, to keep the overall measurement time feasible.
Figure~\ref{fg_grading_template_example} shows an excerpt of the spreadsheet.
The spreadsheet includes a column for each measured software package. \citep{Dong2021-Data} provides the full set of measurement data.

@@

The full template consists of 108 questions categorized under 9 qualities.  We
designed the questions to be unambiguous, quantifiable, and measurable with
limited time and domain knowledge. We group the measures under headings for each
quality, and one for summary information. The summary information (shown in
Figure~\ref{fg_grading_template_example}) is the first section of the template.
This section summarizes general information, such as the software name, purpose,
platform, programming language, publications about the software, the first
release and the most recent change date, website, source code repository of the
product, number of developers, etc.  We only measure open source software that
is ready for public use by other researchers (following the definition of the
public category from \citep{GewaltigAndCannon2012}). Information in the summary
section sets the context for the project, but it does not directly affect the
grading scores.

For measuring each quality, we ask several questions and the typical answers are
among the collection of ``yes'', ``no'', ``n/a'', ``unclear'', a number, a
string, a date, a set of strings, etc. The grader assigns each quality an
overall score, between 1 and 10, based on all the previous questions.  Several
of the qualities use the word ``surface''.  This is to highlight that, for these
qualities in particular, the best that we can do is a shallow measure.  For
instance, we are not currently doing any experiments to measure usability.
Instead, we are looking for an indication that the developers considered
usability.  We do this by looking for cues in the documentation, like a getting
started manual, a user manual and a statement of expected user characteristics.
We use two freeware tools to collect repository related data:
\href{https://github.com/tomgi/git_stats}{GitStats} \citep{Gieniusz2019} and
\href{https://github.com/boyter/scc}{Sloc Cloc and Code (scc)}
\citep{Boyter2021}.  Details on how we measure each quality are provided in
\citet{SmithEtAl2024_MI_SOP}.

@@

Saaty developed AHP in the 1970s, and people have widely used it since to make
and analyze multiple criteria decisions \citep{VaidyaEtAl2006}. AHP organizes
multiple criteria in a hierarchical structure and uses pairwise comparisons
between alternatives to calculate relative ratios \citep{Saaty1990}. AHP works
with sets of $n$ \textit{options} and $m$ \textit{criteria}.  In our project
$n=29$ and $m=9$ since there are 29 options (software products) and 9 criteria
(qualities). With AHP the sum of the grades (scores) for all products for a
given quality will be 1.0.  We rank the software for each of the qualities, and
then we combine the quality rankings into an overall ranking based on the
relative priorities between qualities.

@@

In this section we review the selected software packages.  We used a two-step
process for selecting the software packages: i) identify software candidates in
the chosen domain; and, ii) filter the list to remove less relevant members
\citep{SmithEtAl2021}.

We initially identified 48 MI candidate software projects from the literature
\citep{Bjorn2017, Bruhschwein2019, Haak2015}, on-line articles \citep{Emms2019,
Hasan2020, Mu2019}, and forum discussions \citep{Samala2014}.  The full list of
48 packages is available in \citet{Dong2021}.  To reduce the length of the list
to a manageable number (29 in this case), we filtered the original list as
follows:

@@

Table~\ref{tab_final_list} shows the 29 software packages that we measured,
along with summary data collected in the year 2020. We arrange the items in
descending order of LOC. We found the initial release dates (Rlsd) for most
projects and marked the two unknown dates with ``?''. The date of the last
update is the date of the latest update, at the time of measurement. We found
funding information (Fnd) for only eight projects.  For the Number Of
Contributors (NOC) we considered anyone who made at least one accepted commit as
a contributor. The NOC is not usually the same as the number of long-term
project members, since many projects received change requests and code from the
community.  With respect to the OS, 25 packages work on all three OSs: Windows
(W), macOS (M), and Linux (L). Although the usual approach to cross-platform
compatibility was to work natively on multiple OSes, five projects achieved
platform-independence via web applications. The full measurement data for all
packages is available in \citet{Dong2021-Data}.

\begin{table}[!ht]
\centering
\begin{tabular}{p{6cm}lllllllll}
\toprule
\multirow{2}{*}{Software} & \multirow{2}{*}{Rlsd} & \multirow{2}{*}{Updated} & \multirow{2}{*}{Fnd} & \multirow{2}{*}{NOC} & \multirow{2}{*}{LOC} & \multicolumn{3}{c}{OS} & \multirow{2}{*}{Web} \\ \cline{7-9}
 &  &  &  &  &  & W & M & L &  \\ \midrule
ParaView \citep{Ahrens2005} & 2002 & 2020-10 & \checkmark & 100 & 886326 & \checkmark & \checkmark & \checkmark & \checkmark \\
Gwyddion \citep{Nevcas2012} & 2004 & 2020-11 &  & 38 & 643427 & \checkmark & \checkmark & \checkmark &  \\
Horos \citep{horosproject2020} & ? & 2020-04 &  & 21 & 561617 &  & \checkmark &  &  \\
OsiriX Lite \citep{PixmeoSARL2019} & 2004 & 2019-11 &  & 9 & 544304 &  & \checkmark &  &  \\
3D Slicer \citep{Kikinis2014} & 1998 & 2020-08 & \checkmark & 100 & 501451 & \checkmark & \checkmark & \checkmark &  \\
Drishti \citep{Limaye2012} & 2012 & 2020-08 &  & 1 & 268168 & \checkmark & \checkmark & \checkmark &  \\
Ginkgo CADx \citep{Wollny2020} & 2010 & 2019-05 &  & 3 & 257144 & \checkmark & \checkmark & \checkmark &  \\
GATE \citep{Jan2004} & 2011 & 2020-10 &  & 45 & 207122 &  & \checkmark & \checkmark &  \\
3DimViewer \citep{TESCAN2020} & ? & 2020-03 & \checkmark & 3 & 178065 & \checkmark & \checkmark &  &  \\
medInria \citep{Fillard2012} & 2009 & 2020-11 &  & 21 & 148924 & \checkmark & \checkmark & \checkmark &  \\
BioImage Suite Web \citep{Papademetris2005} & 2018 & 2020-10 & \checkmark & 13 & 139699 &
\checkmark & \checkmark & \checkmark & \checkmark \\
Weasis \citep{Roduit2021} & 2010 & 2020-08 &  & 8 & 123272 & \checkmark & \checkmark & \checkmark &  \\
AMIDE \citep{Loening2017} & 2006 & 2017-01 &  & 4 & 102827 & \checkmark & \checkmark & \checkmark &  \\
XMedCon \citep{Nolf2003} & 2000 & 2020-08 &  & 2 & 96767 & \checkmark & \checkmark & \checkmark &  \\
ITK-SNAP \citep{Yushkevich2006} & 2006 & 2020-06 & \checkmark & 13 & 88530 & \checkmark & \checkmark & \checkmark &  \\
Papaya \citep{UTHSCSA2019} & 2012 & 2019-05 &  & 9 & 71831 & \checkmark & \checkmark & \checkmark &  \\
OHIF Viewer \citep{Ziegler2020} & 2015 & 2020-10 &  & 76 & 63951 & \checkmark & \checkmark & \checkmark & \checkmark \\
SMILI \citep{Chandra2018} & 2014 & 2020-06 &  & 9 & 62626 & \checkmark & \checkmark & \checkmark &  \\
INVESALIUS 3 \citep{Amorim2015} & 2009 & 2020-09 &  & 10 & 48605 & \checkmark & \checkmark & \checkmark &  \\
dwv \citep{Martelli2021} & 2012 & 2020-09 &  & 22 & 47815 & \checkmark & \checkmark & \checkmark & \checkmark \\
DICOM Viewer \citep{Afsar2021} & 2018 & 2020-04 & \checkmark & 5 & 30761 & \checkmark & \checkmark & \checkmark &  \\
MicroView \citep{ParallaxInnovations2020} & 2015 & 2020-08 &  & 2 & 27470 & \checkmark & \checkmark & \checkmark &  \\
MatrixUser \citep{Liu2016} & 2013 & 2018-07 &  & 1 & 23121 & \checkmark & \checkmark & \checkmark &  \\
Slice:Drop \citep{Haehn2013} & 2012 & 2020-04 &  & 3 & 19020 & \checkmark & \checkmark & \checkmark & \checkmark \\
dicompyler \citep{Panchal2010} & 2009 & 2020-01 &  & 2 & 15941 & \checkmark & \checkmark &  &  \\
Fiji \citep{Schindelin2012} & 2011 & 2020-08 & \checkmark & 55 & 10833 & \checkmark & \checkmark & \checkmark &  \\
ImageJ \citep{Rueden2017} & 1997 & 2020-08 & \checkmark & 18 & 9681 & \checkmark & \checkmark & \checkmark &  \\
MRIcroGL \citep{Rorden2021} & 2015 & 2020-08 &  & 2 & 8493 & \checkmark & \checkmark & \checkmark &  \\
DicomBrowser \citep{Archie2012} & 2012 & 2020-08 &  & 3 & 5505 & \checkmark & \checkmark & \checkmark &  \\ \bottomrule
\end{tabular}
\caption{Final software list (sorted in descending order of the number of Lines
Of Code (LOC))}
\label{tab_final_list}
\end{table}

@@

Figure~\ref{fg_correctness_verifiability_scores} shows the scores of correctness
and verifiability. Generally speaking, the packages with higher scores adopted
more techniques to improve correctness, and had better documentation for us to
verify against.  For instance, we looked for evidence of unit testing, since it
benefits most parts of the software's life cycle, such as designing, coding,
debugging, and optimization \citep{Hamill2004}.  We only found evidence of unit
testing for about half of the projects. We identified five projects using CI/CD
tools: \textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji}, \textit{dwv}, and
\textit{OHIF Viewer}.

@@

Digital Imaging and Communications in Medicine (DICOM) ``defines the formats for
medical images that can be exchanged with the data and quality necessary for
clinical use'' \citep{MITA2021}. According to their documentation, all 29
software packages should support the DICOM standard. To test robustness, we
prepared two types of image files: correct and incorrect formats (with the
incorrect format created by relabelled a text file to have the ``.dcm''
extension).  All software packages loaded the correct format image, except for
\textit{GATE}, which failed for unknown reasons.  For the broken format,
\textit{MatrixUser}, \textit{dwv}, and \textit{Slice:Drop} ignored the incorrect
format of the file and loaded it regardless. They did not show any error message
and displayed a blank image. \textit{MRIcroGL} behaved similarly except that it
showed a meaningless image. \textit{Drishti} successfully detected the broken
format of the file, but the software crashed as a result.

@@

In Section~\ref{Sec_VsCommunityRanking}, we compare our ranking to the MI
software community's ranking, as a way to judge our results, and as a way to
understand what the MI community currently values. To understand the state of
the practice for MI software relative to other software, we compare MI software
to other research software in Section~\ref{Sec_CompareArtifacts}.  The
comparison is based on the artifacts our measurement exercise found in the MI
repositories, versus what is typically recommended for research software.  We
discuss threats to the validity of our data and conclusions in section
(Section~\ref{sec_threats_to_validity}).

@@

Further reason for discrepancies between our ranking and the community's ranking
is that we weighted all qualities equally. This is not likely how users
implicitly rank the different qualities. As a result, some projects with high
community popularity may have scored lower with our method because of a
relatively higher (compared to the scientific community's implicit ranking)
weighting of the poor scores for some qualities. A further explanation for
discrepancies between our measures and the star measures may also be due to
inaccuracy with using stars to approximate popularity.  Stars are not an ideal
measure because stars represent the community's feeling in the past more than
they measure current preferences \citep{Szulik2017}.  The issue with stars is
that they tend only to be added, not removed.  A final reason for
inconsistencies between our ranking and the community's ranking is that, as for
consumer products, more factors influence popularity than just quality.

@@

As part of filling in the measurement template (from
Section~\ref{sec_grading_software}), we summarized the artifacts observed in
each MI package. Table~\ref{artifactspresent} groups the artifacts by frequency
into categories of common (20 to 29 ($>$67\%) packages), uncommon (10 to 19
(33-67\%) packages), and rare (1 to 9 ($<$33\%) packages). \citet{Dong2021-Data}
summarizes the full measurements.  Tables~\ref{tab_maintainability_docs}
and~\ref{tab_Visibility/Transparency_docs} show the details on which projects
use which types of artifacts for documents related to maintainability and
visibility, respectively.

@@

\item \textbf{A Contributing file} provides new contributors with the information
that they need to start adding/modifying the repository's files.
\citet{Abdalla2016} provides a simple template for creating an open-source
contributor guideline.

>>>>

%\usepackage[round, sort]{natbib}

@@

% prepared for: Journal of Imaging Informatics in Medicine
% https://link.springer.com/journal/10278

@@

We aim to study the state of software development practice for Medical Imaging
(MI) software using data available in open source repositories.  MI tools use
images of the interior of the body (from sources such as Magnetic Resonance
Imaging (MRI), Computed Tomography (CT), Positron Emission Tomography (PET) and
Ultrasound) to provide information for diagnostic, analytic, and medical
applications \cite{FDA2021, enwiki:1034887445, Zhang2008}.
Figure~\ref{Fig_Example}, which shows an image of the brain, highlights the
importance and value of MI. Through MI medical practitioners and researchers can
noninvasively gain insights into the human body, including information on
injuries and illnesses. Given the importance of MI software and the high number
of competing software projects, we wish to understand the merits and drawbacks
of the current development processes, tools, and methodologies. We aim to assess
through a software engineering lens the quality of the existing software with
the hope of highlighting standout examples, and providing guidelines and
recommendations for future development.

@@

Not only do we wish to gain insight into the state of the practice for MI
software, we also wish to understand the impact of the often cited gap, or
chasm, between software engineering and research software \cite{Kelly2007,
Storer2017}. Although scientists spend a substantial proportion of their working
hours on software development \cite{Hannay2009, Prabhu2011}, many developers
learn software engineering skills by themselves or from their peers, instead of
from proper training \cite{Hannay2009}. In the past many scientists have shown
ignorance and indifference to standard software engineering concepts
\cite{Hannay2009}. For instance, according to a survey by Prabhu in 2011
\cite{Prabhu2011}, more than half of their 114 subjects did not use a proper
debugger when coding.

To gain insight, we devised 10 research questions, which can be applied to MI,
as well as to other domains, of research software \cite{SmithEtAl2021,
SmithAndMichalski2022}.  We designed the questions to learn about the
community's interest in, and experience with, software artifacts, tools,
principles, processes, methodologies, and qualities.  When we mention artifacts
we mean the documents, scripts and code that constitutes a software development
project. Example artifacts include requirements, specifications, user manuals,
unit tests, system tests, usability tests, build scripts, API (Application
Programming Interface) documentation, READMEs, license documents, process
documents, and code.  Once we have learned what MI developers do, we then put
this information in context by contrasting MI software against the trends shown
by developers in other research software communities.

@@

To make the project feasible, we only cover MI visualization software.  As a
consequence we are excluding many other categories of MI software, including
Segmentation, Registration, Visualization, Enhancement, Quantification,
Simulation, plus MI archiving and telemedicine systems (Compression, Storage,
and Communication) \cite{Bankman2000, Angenent2006}.  We also exclude
Statistical Analysis and Image-based Physiological Modelling
\cite{enwiki:1034877594} and Feature Extraction, Classification, and
Interpretation \cite{Kim2011}. Software that provides MI support functions is
also out of scope; therefore, we have not assessed the toolkit libraries VTK
\cite{SchroederEtAl2006} and ITK \cite{McCormick2014}.  Finally, Picture
Archiving and Communication System (PACS), which helps users to economically
store and conveniently access images \cite{Choplin1992}, are considered out of
scope.

@@

We designed a general methodology to assess the state of the practice for
research software \cite{SmithEtAl2021, SmithAndMichalski2022}.  Our methodology
has been applied to MI software \cite{Dong2021} and Lattice Boltzmann Solvers
\cite{Michalski2021, SmithEtAl2024}.  This methodology builds off prior work to
assess the state of the practice for such domains as Geographic Information
Systems \cite{smith2018state}, Mesh Generators \cite{smith2016state},
Seismology software \cite{Smith2018Seismology}, and Statistical software for
psychology \cite{smith2018statistical}.  In keeping with the previous
methodology, we have maintained the constraint that the work load for measuring
a given domain should be feasible for a team as small as one person, and for a
short time, ideally around a person month of effort. We consider a person month
as $20$ working days ($4$ weeks in a month, with $5$ days of work per week) at
$8$ person hours per day, or $20 \times 8 = 160$ person hours.

With our methodology, we first choose a research software domain (in the current
case MI) and identify a list of about 30 software packages. (For measuring MI we
used 29 software packages.)  The selected packages are vetted by a domain
expert. We approximately measure the qualities of each package by filling in a
grading template. The grader looks at the contents of the open source repository
and repository based metrics, such as the number of files, number of lines of
code, percentage of issues that are closed, etc.  With the quantitative data in
the grading template, we rank the software with the Analytic Hierarchy Process
(AHP). Further details on interaction with the domain expert, software
qualities, grading the software and AHP are found below and in
\cite{SmithEtAl2024_MI_SOP}.

@@

Quality is defined as a measure of the excellence or worth of an entity.  As is
common practice, we do not think of quality as a single measure, but rather as a
set of measures.  That is, quality is a collection of different qualities, often
called ``ilities.''  For this study we selected 9 qualities to measure:
installability, correctness/ verifiability, reliability, robustness, usability,
maintainability, reusability, understandability, and visibility/transparency.
With the exception of installability, all the qualities are defined in Ghezzi et
al. (2003) \cite{GhezziEtAl2003}. Installability is defined as the effort
required for the installation and/or uninstallation of software in a specified
environment \cite{ISO/IEC25010, lenhard2013measuring}.

@@

We grade the selected software using an existing measurement template
\cite{SmithEtAl2021}.  The template provides measures of the qualities listed in
Section~\ref{sec_software_quality}.  For each software package, we fill in the
template questions. To stay within the target of 160 person hours to measure the
domain, we allocated between one and four hours for each package. Project
developers can be contacted for help regarding installation, if necessary, but
we impose a cap of about two hours on the installation process, to keep the
overall measurement time feasible.  Figure~\ref{fg_grading_template_example}
shows an excerpt of the spreadsheet.  The spreadsheet includes a column for each
measured software package. \cite{Dong2021-Data} provides the full set of
measurement data.

@@

The full template consists of 108 questions categorized under 9 qualities.  We
designed the questions to be unambiguous, quantifiable, and measurable with
limited time and domain knowledge. We group the measures under headings for each
quality, and one for summary information. The summary information (shown in
Figure~\ref{fg_grading_template_example}) is the first section of the template.
This section summarizes general information, such as the software name, purpose,
platform, programming language, publications about the software, the first
release and the most recent change date, website, source code repository of the
product, number of developers, etc.  We only measure open source software that
is ready for public use by other researchers (following the definition of the
public category from \cite{GewaltigAndCannon2012}). Information in the summary
section sets the context for the project, but it does not directly affect the
grading scores.

For measuring each quality, we ask several questions and the typical answers are
among the collection of ``yes'', ``no'', ``n/a'', ``unclear'', a number, a
string, a date, a set of strings, etc. The grader assigns each quality an
overall score, between 1 and 10, based on all the previous questions.  Several
of the qualities use the word ``surface''.  This is to highlight that, for these
qualities in particular, the best that we can do is a shallow measure.  For
instance, we are not currently doing any experiments to measure usability.
Instead, we are looking for an indication that the developers considered
usability.  We do this by looking for cues in the documentation, like a getting
started manual, a user manual and a statement of expected user characteristics.
We use two freeware tools to collect repository related data:
\href{https://github.com/tomgi/git_stats}{GitStats} \cite{Gieniusz2019} and
\href{https://github.com/boyter/scc}{Sloc Cloc and Code (scc)}
\cite{Boyter2021}.  Details on how we measure each quality are provided in
Smith et al.\ (2024) \cite{SmithEtAl2024_MI_SOP}.

@@

Saaty developed AHP in the 1970s, and people have widely used it since to make
and analyze multiple criteria decisions \cite{VaidyaEtAl2006}. AHP organizes
multiple criteria in a hierarchical structure and uses pairwise comparisons
between alternatives to calculate relative ratios \cite{Saaty1990}. AHP works
with sets of $n$ \textit{options} and $m$ \textit{criteria}.  In our project
$n=29$ and $m=9$ since there are 29 options (software products) and 9 criteria
(qualities). With AHP the sum of the grades (scores) for all products for a
given quality will be 1.0.  We rank the software for each of the qualities, and
then we combine the quality rankings into an overall ranking based on the
relative priorities between qualities.

@@

In this section we review the selected software packages.  We used a two-step
process for selecting the software packages: i) identify software candidates in
the chosen domain; and, ii) filter the list to remove less relevant members
\cite{SmithEtAl2021}.

We initially identified 48 MI candidate software projects \cite{Dong2021} from
the literature \cite{Bjorn2017, Bruhschwein2019, Haak2015}, on-line articles
\cite{Emms2019, Hasan2020, Mu2019}, and forum discussions \cite{Samala2014}.  To
reduce the length of the list to a manageable number (29 in this case), we
filtered the original list as follows:

@@

Table~\ref{tab_final_list} shows the 29 software packages that we measured,
along with summary data collected in the year 2020. We arrange the items in
descending order of LOC. We found the initial release dates (Rlsd) for most
projects and marked the two unknown dates with ``?''. The date of the last
update is the date of the latest update, at the time of measurement. We found
funding information (Fnd) for only eight projects.  For the Number Of
Contributors (NOC) we considered anyone who made at least one accepted commit as
a contributor. The NOC is not usually the same as the number of long-term
project members, since many projects received change requests and code from the
community.  With respect to the OS, 25 packages work on all three OSs: Windows
(W), macOS (M), and Linux (L). Although the usual approach to cross-platform
compatibility was to work natively on multiple OSes, five projects achieved
platform-independence via web applications. The full measurement data for all
packages is available on \href{https://data.mendeley.com/datasets/k3pcdvdzj2/1}
{Mendeley Data}.

\begin{table}[!ht]
\centering
\begin{tabular}{p{6cm}lllllllll}
\toprule
\multirow{2}{*}{Software} & \multirow{2}{*}{Rlsd} & \multirow{2}{*}{Updated} & \multirow{2}{*}{Fnd} & \multirow{2}{*}{NOC} & \multirow{2}{*}{LOC} & \multicolumn{3}{c}{OS} & \multirow{2}{*}{Web} \\ \cline{7-9}
 &  &  &  &  &  & W & M & L &  \\ \midrule
ParaView \cite{Ahrens2005} & 2002 & 2020-10 & \checkmark & 100 & 886326 & \checkmark & \checkmark & \checkmark & \checkmark \\
Gwyddion \cite{Nevcas2012} & 2004 & 2020-11 &  & 38 & 643427 & \checkmark & \checkmark & \checkmark &  \\
Horos \cite{horosproject2020} & ? & 2020-04 &  & 21 & 561617 &  & \checkmark &  &  \\
OsiriX Lite \cite{PixmeoSARL2019} & 2004 & 2019-11 &  & 9 & 544304 &  & \checkmark &  &  \\
3D Slicer \cite{Kikinis2014} & 1998 & 2020-08 & \checkmark & 100 & 501451 & \checkmark & \checkmark & \checkmark &  \\
Drishti \cite{Limaye2012} & 2012 & 2020-08 &  & 1 & 268168 & \checkmark & \checkmark & \checkmark &  \\
Ginkgo CADx \cite{Wollny2020} & 2010 & 2019-05 &  & 3 & 257144 & \checkmark & \checkmark & \checkmark &  \\
GATE \cite{Jan2004} & 2011 & 2020-10 &  & 45 & 207122 &  & \checkmark & \checkmark &  \\
3DimViewer \cite{TESCAN2020} & ? & 2020-03 & \checkmark & 3 & 178065 & \checkmark & \checkmark &  &  \\
medInria \cite{Fillard2012} & 2009 & 2020-11 &  & 21 & 148924 & \checkmark & \checkmark & \checkmark &  \\
BioImage Suite Web \cite{Papademetris2005} & 2018 & 2020-10 & \checkmark & 13 & 139699 &
\checkmark & \checkmark & \checkmark & \checkmark \\
Weasis \cite{Roduit2021} & 2010 & 2020-08 &  & 8 & 123272 & \checkmark & \checkmark & \checkmark &  \\
AMIDE \cite{Loening2017} & 2006 & 2017-01 &  & 4 & 102827 & \checkmark & \checkmark & \checkmark &  \\
XMedCon \cite{Nolf2003} & 2000 & 2020-08 &  & 2 & 96767 & \checkmark & \checkmark & \checkmark &  \\
ITK-SNAP \cite{Yushkevich2006} & 2006 & 2020-06 & \checkmark & 13 & 88530 & \checkmark & \checkmark & \checkmark &  \\
Papaya \cite{UTHSCSA2019} & 2012 & 2019-05 &  & 9 & 71831 & \checkmark & \checkmark & \checkmark &  \\
OHIF Viewer \cite{Ziegler2020} & 2015 & 2020-10 &  & 76 & 63951 & \checkmark & \checkmark & \checkmark & \checkmark \\
SMILI \cite{Chandra2018} & 2014 & 2020-06 &  & 9 & 62626 & \checkmark & \checkmark & \checkmark &  \\
INVESALIUS 3 \cite{Amorim2015} & 2009 & 2020-09 &  & 10 & 48605 & \checkmark & \checkmark & \checkmark &  \\
dwv \cite{Martelli2021} & 2012 & 2020-09 &  & 22 & 47815 & \checkmark & \checkmark & \checkmark & \checkmark \\
DICOM Viewer \cite{Afsar2021} & 2018 & 2020-04 & \checkmark & 5 & 30761 & \checkmark & \checkmark & \checkmark &  \\
MicroView \cite{ParallaxInnovations2020} & 2015 & 2020-08 &  & 2 & 27470 & \checkmark & \checkmark & \checkmark &  \\
MatrixUser \cite{Liu2016} & 2013 & 2018-07 &  & 1 & 23121 & \checkmark & \checkmark & \checkmark &  \\
Slice:Drop \cite{Haehn2013} & 2012 & 2020-04 &  & 3 & 19020 & \checkmark & \checkmark & \checkmark & \checkmark \\
dicompyler \cite{Panchal2010} & 2009 & 2020-01 &  & 2 & 15941 & \checkmark & \checkmark &  &  \\
Fiji \cite{Schindelin2012} & 2011 & 2020-08 & \checkmark & 55 & 10833 & \checkmark & \checkmark & \checkmark &  \\
ImageJ \cite{Rueden2017} & 1997 & 2020-08 & \checkmark & 18 & 9681 & \checkmark & \checkmark & \checkmark &  \\
MRIcroGL \cite{Rorden2021} & 2015 & 2020-08 &  & 2 & 8493 & \checkmark & \checkmark & \checkmark &  \\
DicomBrowser \cite{Archie2012} & 2012 & 2020-08 &  & 3 & 5505 & \checkmark & \checkmark & \checkmark &  \\ \bottomrule
\end{tabular}
\caption{Final software list (sorted in descending order of the number of Lines
Of Code (LOC))}
\label{tab_final_list}
\end{table}

@@

Figure~\ref{fg_correctness_verifiability_scores} shows the scores of correctness
and verifiability. Generally speaking, the packages with higher scores adopted
more techniques to improve correctness, and had better documentation for us to
verify against.  For instance, we looked for evidence of unit testing, since it
benefits most parts of the software's life cycle, such as designing, coding,
debugging, and optimization \cite{Hamill2004}.  We only found evidence of unit
testing for about half of the projects. We identified five projects using CI/CD
tools: \textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji}, \textit{dwv}, and
\textit{OHIF Viewer}.

@@

Digital Imaging and Communications in Medicine (DICOM) ``defines the formats for
medical images that can be exchanged with the data and quality necessary for
clinical use'' \cite{MITA2021}. According to their documentation, all 29
software packages should support the DICOM standard. To test robustness, we
prepared two types of image files: correct and incorrect formats (with the
incorrect format created by relabelled a text file to have the ``.dcm''
extension).  All software packages loaded the correct format image, except for
\textit{GATE}, which failed for unknown reasons.  For the broken format,
\textit{MatrixUser}, \textit{dwv}, and \textit{Slice:Drop} ignored the incorrect
format of the file and loaded it regardless. They did not show any error message
and displayed a blank image. \textit{MRIcroGL} behaved similarly except that it
showed a meaningless image. \textit{Drishti} successfully detected the broken
format of the file, but the software crashed as a result.

@@

In Section~\ref{Sec_VsCommunityRanking}, we compare our ranking to the MI
software community's ranking, as a way to judge our results, and as a way to
understand what the MI community currently values. To understand the state of
the practice for MI software relative to other software, we compare MI software
to other research software in Section~\ref{Sec_CompareArtifacts}.  The
comparison is based on the artifacts our measurement exercise found in the MI
repositories, versus what is typically recommended for research software.  We
provide details on artifacts that are recommended but rarely observed for MI
software. Section~\ref{sec_threats_to_validity} presents threats to the validity
of our data and conclusions.

@@

Further reason for discrepancies between our ranking and the community's ranking
is that we weighted all qualities equally. This is not likely how users
implicitly rank the different qualities. As a result, some projects with high
community popularity may have scored lower with our method because of a
relatively higher (compared to the scientific community's implicit ranking)
weighting of the poor scores for some qualities. A further explanation for
discrepancies between our measures and the star measures may also be due to
inaccuracy with using stars to approximate popularity.  Stars are not an ideal
measure because stars represent the community's feeling in the past more than
they measure current preferences \cite{Szulik2017}.  The issue with stars is
that they tend only to be added, not removed.  A final reason for
inconsistencies between our ranking and the community's ranking is that, as for
consumer products, more factors influence popularity than just quality.

@@

As part of filling in the measurement template (from
Section~\ref{sec_grading_software}), we summarized the artifacts observed in
each MI package. Table~\ref{artifactspresent} groups the artifacts by frequency
into categories of common (20 to 29 ($>$67\%) packages), uncommon (10 to 19
(33-67\%) packages), and rare (1 to 9 ($<$33\%) packages).
Tables~\ref{tab_maintainability_docs} and~\ref{tab_Visibility/Transparency_docs}
show the details on which projects use which types of artifacts for documents
related to maintainability and visibility, respectively.

@@

\item \textbf{A Contributing file} provides new contributors with the
information that they need to start adding/modifying the repository's files.
Abdalla \cite{Abdalla2016} provides a simple template for creating an
open-source contributor guideline.
